---
title: "REF IR/Political Science Prediction Models (version 0.3)"
author: "Christopher Gandrud"
date: "25 February 2016"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document:
    fig_caption: yes
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(rio)
library(dplyr)
library(randomForest)
library(DataCombine)
library(ggplot2)
library(gridExtra)
library(tidyr)

set.seed(200)

main <- import('data/gpa_impact_google.csv')
main <- main %>% DropNA('ref_gpa')

main$school_label <- 'Other'
main$school_label[main$UKPRN == 10001478] <- 'City'
main$school_label[main$UKPRN == 10004063] <- 'LSE'
main$school_label[main$UKPRN == 10007774] <- 'Oxford'
main$school_label[main$UKPRN == 10007791] <- 'Essex'

main$highlight <- 0
main$highlight[main$school_label != 'Other'] <- 1
main$highlight <- main$highlight %>% as.factor

# Create training and test sets
rownames(main) <- 1:nrow(main)
rows <- sample(x = 1:nrow(main), size = 0.7 * nrow(main))

train <- main[rows,]
test <- main[! rownames(main) %in% rows,]

# Estimate model
m1 <- randomForest(ref_gpa ~ google_40_perc + mean_impact + top_up_perc, data = train)

varImpPlot(m1)

# Compare predicted to actual in the test data set
pred <- predict(object = m1, newdata = test)
compare_1 <- data.frame(actual = test$ref_gpa, predicted = pred)
compare_1$residuals <- compare_1$actual - compare_1$predicted
```

I conducted a simple Random Forest Regression to examine how IR/Political Science REF 2014 Output GPAs could be predicted using:

- **Mean Journal Impact Factor** for all of the journal article submissions that each university made. Note: if a journal was not assigned an impact factor[^impact_caveat] it is effectively given an impact factor of 0.

- **Percent of journal article** submissions from journals in the **top 20** IR or Political Science categories assembled by Google Scholar.

- **Percent of non-edited books** submitted that were published by a **top university press** (see table in the Appendix for the complete list).

All of these metrics are highly correlated with REF Output GPAs. Mean impact factor has a correlation coefficient of `r cor(main$ref_gpa, main$mean_impact) %>% round(digits = 2)` with REF Output GPAs. The REF GPA correlation with the Google Scholar metric is `r cor(main$ref_gpa, main$google_40_perc) %>% round(digits = 2)` and `r cor(main$ref_gpa, main$top_up_perc) %>% round(digits = 2)` with the percentage of books submitted that were published by a top university press. The following figure further illustrates these close relationships and City University's placement within them. 

It is important to note that neither of these metrics contain information on other materials including edited volues which are also submitted to the REF.

```{r descript, echo=FALSE, warning=FALSE, fig.cap='Comparing Universities\' REF 2014 Output GPA to Journal Research Output Metrics', fig.lp='fig:'}

custom_col <- c("#F2300F", "#E1BD6D", "#0b775e", "#EABE94", "#35274A")

gathered <- main %>% gather(metric, value, c(3:4, 7))
gathered$metric[gathered$metric == 'google_40_perc'] <- '% at Top of Google'
gathered$metric[gathered$metric == 'mean_impact'] <- 'Mean Impact Factor'
gathered$metric[gathered$metric == 'top_up_perc'] <- '% of Books in Top UP'

ggplot(gathered, aes(value, ref_gpa)) +
    stat_smooth(colour = 'black') +
    geom_point(aes(colour = school_label, size = highlight, alpha = highlight)) +
    facet_wrap(~ metric, scales = 'free', ncol = 2) +
    scale_colour_manual(values = custom_col, name = '') +
    scale_alpha_discrete(range = c(0.3, 1)) +
    #xlab('\nMean Journal Submission Impact Factor') +
    ylab('2014 REF Outputs GPA\n') +
    xlab('') +
    theme_bw() + guides(size = FALSE, alpha = FALSE)
```

# More Complex Model: Google Scholar + Impact Factor + Books

To examine how well these metrics predict REF Output GPAs I first ran the random forest regression model on a random sample of 70% of the 56 universities (i.e. 38) that made REF submissions for IR/Political Science. I then used the estimates from the model to predict the REF Output scores of the remaining 30% (i.e. 17 universities). The following figure compares the actual REF GPA scores to the predictions. Note: if the model perfectly predicted the GPA score then each dot would lie one the 45 degree line. The mean absolute prediction error when using the two journal metrics was `r mean(abs(compare_1$residuals)) %>% round(digits = 2)`. In other words, on average the model incorrectly predicted the REF GPA score by `r mean(abs(compare_1$residuals)) %>% round(digits = 2)` GPA points or `r (mean(abs(compare_1$residuals))  / 4) %>% round(digits = 3) * 100`% of the GPA scale.

```{r fig.cap='Actual vs. Predicted 2014 REF Output GPAs Using All Output Metrics for a Test Set of 17 Randomly Selected Universities'}
ggplot(compare_1, aes(actual, predicted)) +
    geom_point() +
    scale_y_continuous(limits = c(0, 4)) +
    scale_x_continuous(limits = c(0, 4)) +
    geom_abline(intercept = 0, slope = 1, linetype = 'dotted') +
    xlab('\nActual REF GPA') + ylab('Predicted REF GPA\n') +
    theme_bw()
```

# Simpler model: Google Scholar Only

```{r include=FALSE}
# Google only
m2 <- randomForest(ref_gpa ~ google_40_perc, data = train)

# Compare predicted to actual in the test data set
pred_2 <- predict(object = m2, newdata = test)
compare_2 <- data.frame(actual = test$ref_gpa, predicted = pred_2)
compare_2$residuals <- compare_2$actual - compare_2$predicted

# Google Plus
m3 <- randomForest(ref_gpa ~ google_40_plus, data = train)

# Compare predicted to actual in the test data set
pred_3 <- predict(object = m3, newdata = test)
compare_3 <- data.frame(actual = test$ref_gpa, predicted = pred_3)
compare_3$residuals <- compare_3$actual - compare_3$predicted
```

The percentage of journal submissions in the top Google Scholar lists is more strongly correlated with REF GPA scores than impact factors. Would a simpler model using just the Google Scholar metric perform just as well as the more complex two metric model? The following figures shows actual vs. predicted GPA scores for this model. The mean absolute prediction error when using only the Google Scholar metric was `r mean(abs(compare_2$residuals)) %>% round(digits = 3)`. In other words, on average the model incorrectly predicted the REF GPA score by `r mean(abs(compare_2$residuals)) %>% round(digits = 2)` GPA points or `r (mean(abs(compare_2$residuals))  / 4) %>% round(digits = 3) * 100`% of the GPA scale. The Google Scholar Only model slightly outperforms the more complex model that also included information on impact factors.

```{r fig.cap='Actual vs. Predicted 2014 REF Output GPAs Using Google Scholar Metric for a Test Set of 17 Randomly Selected Universities'}
ggplot(compare_2, aes(actual, predicted)) +
    geom_point() +
    scale_y_continuous(limits = c(0, 4)) +
    scale_x_continuous(limits = c(0, 4)) +
    geom_abline(intercept = 0, slope = 1, linetype = 'dotted') +
    xlab('\nActual REF GPA') + ylab('Predicted REF GPA\n') +
    theme_bw()

```


# Simple, But a Little More Complex: Google Plus

The Google Top 20 IR and Political Science lists are notably lacking important political economy journals, including *Review of International Political Economy* and *New Political Economy*. Does adding these journals to a "Google Scholar Plus" variable improve prediction performance? The following figure shows the predicted vs. actual REF GPAs for our test sample using the Google Scholar Plus variable. The mean absolute prediction error when using only the Google Scholar Plus metric was `r mean(abs(compare_3$residuals)) %>% round(digits = 3)`. In other words, on average the model incorrectly predicted the REF GPA score by `r mean(abs(compare_3$residuals)) %>% round(digits = 3)` GPA points or `r (mean(abs(compare_3$residuals))  / 4) %>% round(digits = 3) * 100`% of the GPA scale. The Google Scholar Plus model slightly outperforms both the Two Metric model and the Google Scholar Only model.

```{r fig.cap='Actual vs. Predicted 2014 REF Output GPAs Using Google Scholar Plus Metric for a Test Set of 17 Randomly Selected Universities'}
ggplot(compare_3, aes(actual, predicted)) +
    geom_point() +
    scale_y_continuous(limits = c(0, 4)) +
    scale_x_continuous(limits = c(0, 4)) +
    geom_abline(intercept = 0, slope = 1, linetype = 'dotted') +
    xlab('\nActual REF GPA') + ylab('Predicted REF GPA\n') +
    theme_bw()

```


# Conclusion

For the International Relations/Political Science 2014 REF we can create highly accurate predictions of Output GPA using only universities' percentage of journal submissions that were from the Google Scholar IR and Political Science top 20 lists. Using subject specific knowledge to select two additional journals that are prominent in international political economy allows us to make even better predictions. It may be possible to make even better predictions by including information on high impact book publishers. 

# Appendix: Top 20 (IR + PS) Google Scholar Journals (February 2016)

For reference, the following is a list of the top 20 journals in the Google Scholar IR and Political Science categories. We also include the most recent Thomson-Reuters Impact factor to enable comparison between the two metrics. Note that the table does not include 40 journals because some journals (*World Politics* and *Journal of Democracy* show up on both the IR and Political Science lists).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Impact factors
impact <- import('data/raw/Impact Factor.xlsx')
impact <- impact[, c(1, 3, 5)]
names(impact) <- c('journal', 'impact_factor', 'google_top')
impact$journal <- impact$journal %>% tolower

impact$journal[impact$journal == 'governance-an international journal of policy administration and institutions'] <- 'governance'
impact$journal[grep('^jcms', impact$journal)] <- 'jcms'
impact$journal[grep('^millennium', impact$journal)] <- 'millennium'


impact$google_top[!is.na(impact$google_top)] <- 1
impact$google_top[is.na(impact$google_top)] <- 0


google_ir <- impact %>% filter(google_top == 1) %>%
                select(-google_top) %>% DropNA('journal')

names(google_ir) <- c('Journal', 'TR Impact Factor')

knitr::kable(google_ir)
```


[^impact_caveat]: E.g. new journals and journals not included in the impact factor list used. Note that we attempted to match all of the submitted articles' journal names with those on the impact factor list. However, due to spelling  variations in the two sets of journal names, some matches may not have been made.

[^HEFCE]: <http://www.dcscience.net/2015_metrictideS2.pdf>
